Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-09-12 paper/paper.tex
add matlab plotting scripts back

\begin{abstract}
We introduce a new approach to compute various properties of independent sets by encoding the problems and the graph into tensor networks with different tensor element algebra.
The type of graph problems that can be computed using this method includes: independence number, the number of independent sets, independence polynomial, maximal independence polynomial, and both counting and enumeration of all maximum independence sets, the large-size independent sets, and all maximal independent sets.
Although the computational complexity inevitably scales exponentially with the graph size, our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant, Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
\end{abstract}

@@

Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
Naive exhaustive search for the MIS requires time $O\left(2^{|V|} \right)$. More efficient exact algorithms include the branching algorithms and dynamic programming.
Sophisticated branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}.
Dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.

The independent set problem is closely related to the clique problem and the vertex cover problem~\cite{Moore2011};
more concretely, an MIS of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
Together, these problems find a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}.
Besides the standard problem of finding the independence number and an MIS, there are many other interesting problems pertaining to independent sets, such as the number of independent sets,
independence polynomial, maximal independence polynomial, and enumeration of MIS or maximal independent set configurations.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions and to identify harder graphs in an ensemble of graphs \red{cite experiment}.

In this work, we introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map different problems into generic tensor network contraction with various tensor element algebra.
Our approach does not necessarily provide a better time complexity compared to dynamic programming,
but the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on CPUs and GPUs, the implementation of which benefits from recent advances in tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021}.
Lastly, we provide a few examples and show that the toolbox can be used to calculate the hard-square entropy constant,
Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.

@@

\begin{example}
$C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is a tensor network representing $C_{ijk} = \sum_{ml}A_{jkm} B_{mil} V_{jm}$,
where $i,j,k$ in the output tensor are open edges (note that $j$ is not an open edge and thus summed over).
Its hypergraph representation is shown below, where we use different colors to annotate different hyperedges.

@@

\section{Generic programming}\label{sec:generic}
In previous works relating tensor networks and combinatoric problems~\cite{Kourtis2019, Biamonte2017},
the elements in the tensor networks are limited to standard number types such as floating point numbers and integers.
Owning to the development of modern compiling technology, we no longer need to limit our imagination to standard number types.
One of the key concepts that push the technology forward is called generic programming:

@@

This definition of generic programming contains two major aspects: a single program for the general setting and efficiency.
To understand the first aspect, suppose we want to write a function that raises an element to a power, $f(x, n) := x^n$.
One can easily write a function for standard number types that computes the power of $x$ in $\log(n)$ steps using the multiply and square trick.
Generic programming does not require $x$ to be a standard number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and a multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as an input, it computes the matrix power without extra efforts.
The second aspect is about performance. For dynamically typed languages such as Python,
one can easily write very general codes, but the efficiency is not guaranteed; for example, the speed of computing the matrix multiplication between two numpy arrays with python objects as elements is much slower than statically typed languages such as C++ and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia takes advantage of just-in-time compilation and multiple dispatch.
When these languages ``see'' a new input type, the compiler can recompile the generic program for the new type.
A myriad of optimizations can be done during the compilation, such as inlining immutable elements with fixed sizes in an array to decrease the cache miss rate when accessing data.
In Julia, these inlined arrays can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

This motivates us to think about what is the most general element type that is allowed in a tensor network contraction program.
We find that as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a semiring with its multiplication operation being commutative, while a semiring is a ring without additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
The requirement of being commutative is for the tensor contraction result to be independent of the contraction order, which could be relaxed in certain cases. In the following sections, we show how to compute the independence polynomial,
the maximal independence polynomial, independence number, the number of independent sets, and enumeration of the MIS and maximal independent set configurations of a general graph $G$ by designing tensor element types as commutative semirings while keeping the tensor network generic~\cite{Stepanov2014}.
Table~\ref{tbl:generictypes} summarizes the problems that can be solved by various tensor element types.

@@

Among the benchmarks shown in \Fig{fig:benchmark}, computing \texttt{max size (CPU)},
\texttt{total size (CPU)}, \texttt{total size (GPU)}, \texttt{IDP (FFT) (CPU)}, \texttt{IDP (FFT) (GPU)} and \texttt{single configuration (bounding) (CPU)} can utilize fast BLAS functions,
hence they are relatively faster comparing to other methods in the same category.
Only algebras requiring dynamic sized structures can not be computed on GPU, here are polynomial algebra for computing independence polynomial and set algebra for enumerating configurations.
When computing the independence polynomial, the finite field approach is the only method without round off error that can run on GPU.
To enumerate the MISs, the technique to bound the enumeration space is important,
without which the largest enumeratible size is only $150$ with 32G memory while being one order slower.

>>>>

\begin{abstract}
We introduce a new approach to compute various properties of independent sets by encoding the problems into tensor networks with different tensor element algebra.
The type of graph problems that can be computed using this method includes: independence number, the number of independent sets, independence polynomial, maximal independence polynomial, and enumeration of the maximum independence set configurations.
Although the computational complexity inevitably scales exponentially with the graph size, the performance of our method using tensor network contraction with generic programming is on par or outperforms state-of-the-art, sophisticated methods;
on the other hand, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant, Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
\end{abstract}

@@

Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
Naive exhaustive search for the MIS requires time $O(2^{|V|})$. More efficient exact algorithms include the branching algorithms and dynamic programming.
Sophisticated branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^n n ^{O(1)}$~\cite{Xiao2017}.
Dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.

The independent set problem is closely related to the clique problem and the vertex cover problem~\cite{Moore2011};
more concretely, an MIS of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
Together, these problems find a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}.
Besides the standard MIS problem, there are many other interesting problems pertaining to independent sets, such as the number of independent sets,
independence polynomial, maximal independence polynomial, and enumeration of optimal and sub-optimal configurations.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions and to identify harder graphs in an ensemble of graphs \red{cite experiment}.

In this work, we introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map different problems into generic tensor network contraction with various tensor element algebra.
Our approach does not necessarily provide a better time complexity compared to dynamic programming,
but the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts.
We benchmark our algorithms, the implementation of which benefits from recent advances in tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021}.
Lastly, we provide a few examples and show that the toolbox can be used to calculate the hard-square entropy constant,
Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.

@@

\begin{example}
$C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is a tensor network representing $C_{ijk} = \sum_{ml}A_{jkm} B_{mil} V_{jm}$,
where $i,j,k$ in the output tensor are open edges.
Its hypergraph representation is shown below, where we use different colors to annotate different hyperedges and open circles to denote open edges.

@@

\section{Generic programming}\label{sec:generic}
In previous works combining tensor network and combinatoric problems~\cite{Kourtis2019,Biamonte2017},
the elements in tensor networks are limited to regular number types such as floating point numbers and integers.
As the development of modern compiling technology, we do not need to limit our imaginations to regular number types anymore.
The key technology playing the role is called the generic programming:

@@

This definition of generic programming contains two major pieces, one is a single program for the general setting, and another is about efficiency.
To understand the first point, suppose we want to write a function to take the power of an element $f(x, n) := x^n$.
One can easily write a function for regular numbers that computes the power of $x$ in $\log(n)$ steps using the multiply and square trick.
A generic programmer does not assume $x$ being a regular number type,
instead it treats $x$ as an element with an associative multiplication operation $\odot$ and multiplicative identity $\mymathbb{1}$.
In such a way, when the program takes a matrix as input, it computes the matrix power without extra efforts.
The second point about performance. For dynamic typed languages like Python,
one can easily write very general codes, however, the efficiency is not guaranteed.
Imagine the speed of computing the matrix multiplication between two numpy arrays with python objects as elements.
The generic programming that we want to discuss is those in static types languages like C++, and Julia~\cite{Bezanson2012}.
C++ uses templates for generic programming while Julia has just in time compiling and multiple dispatch.
When they ``sees'' a new input type, the compiler recompiles the generic program for the new type.
A lot optimization can be done during the compiling, like inlining immutable elements with fixed sizes in an array to decrease the cache miss rate when accessing data.
In Julia, these inlined arrays can even be compiled to GPU devices for faster computation~\cite{Besard2018}.

This leads us to think about what is the most general element type that allowed in a tensor network contraction program.
Our finding is: as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result is contraction order independent.
A commutative semiring is a semiring with its multiplication operation being commutative, while a semiring is a ring without additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
The requirement of being commutative is for the tensor contraction result to be independent of the contraction order, which could be relaxed in certain cases. In the following sections, we show how to compute the independence polynomial,
the maximal independence polynomial, independence number, the number of independent sets, and enumeration of the MIS configurations of a general graph $G$ by designing tensor element types as commutative semirings while keeping the tensor network generic~\cite{Stepanov2014}.
Table~\ref{tbl:generictypes} summarizes the problems that can be solved by various tensor element types.

@@

Among the benchmarks shown in \Fig{fig:benchmark}, computing \texttt{max size (CPU)},
\texttt{total size (CPU)}, \texttt{total size (GPU)}, \texttt{IDP (FFT) (CPU)}, \texttt{IDP (FFT) (GPU)} and \texttt{single configuration (bounding) (CPU)} can utilize fast BLAS functions,
hence they are relatively faster comparing to other methods in the same category.
Only algebras requiring dynamic sized structures can not be computed on GPU, here are polynomial algebra for computing independence polynomial and set algebra for enumerating configurations.
When computing the independence polynomial, the finite field approach is the only method without round off error that can run on GPU.
To enumerate the MISs, the technique to bound the enumeration space is important,
without which the largest enumeratible size is only $150$ with 32G memory while being one order slower.
All these computing times have strong correlation with the tree width,
while the computing times of configurations enumeration also depends on the size of target configuration space.
